---
title: "PS5-Assignment"
author: "Ziyu Lin"
date: "2023-02-14"
output: html_document
---

### 1.Load and check data
#### 1(2).
```{r, message=FALSE}
library(tidyverse)
gapminder <- read_delim("data/gapminder.csv.bz2")
```

- The _gapminder_ data contains `r nrow(gapminder)` rows and `r ncol(gapminder)` columns.

#### 1(3).
```{r}
gapminder %>% 
  sample_n(5)
```

- Yes, it looks OK.

### 2.Descriptive statistics
#### 2(1).
```{r}
iso3_num <- length(unique(gapminder$iso3))
iso2_num <- length(unique(gapminder$iso2))
name_num <- length(unique(gapminder$name))
```
- There are `r iso3_num` countries in iso-3. 
- There are `r iso2_num` countries in iso-2.
- There are `r name_num` countries in names.

#### 2(2)-a.
```{r}
iso2_name <- gapminder %>% 
  group_by(iso2) %>% 
  summarize(n = n_distinct(name)) %>% 
  arrange(desc(n))
iso2_name %>% 
  head(5)
```

- There are one iso-2 codes correspond to more than one name.

#### 2(2)-b
```{r}
iso3_name <- gapminder %>% 
  group_by(name) %>% 
  summarize(n = n_distinct(iso3)) %>% 
  arrange(desc(n))
iso3_name %>% 
  head(5)
```

- There are four countries have more than one iso-3 codes. Two of these entitites are CHANISL and NLD CURACAO.

#### 2(3).
```{r, echo=FALSE}
min(gapminder$time, na.rm = TRUE)
max(gapminder$time, na.rm = TRUE)
```

- The minimum year in this data is `r min(gapminder$time, na.rm = TRUE)`, the maxmum year is `r max(gapminder$time, na.rm = TRUE)`.

### 3.CO2 emissions
#### 3(1).
```{r}
gapminder %>% 
  group_by(time) %>%
  select(co2, co2_PC) %>%
  summarize(volume1 = sum(is.na(co2)),
            volume2 = sum(is.na(co2_PC))) %>%
  arrange(desc(volume1)) %>%
  head(5)
```

- Year 2017, 2018 and 2019 have most missing data.

#### 3(2).
```{r, each=FALSE}
co2_3 <- gapminder %>%
  filter(!is.na(co2)) %>% 
  group_by(time, name) %>% 
  select(name, time, co2) %>%
  filter(name == "United States of America" | name == "China" | name == "India" | name == "Japan" | name == "Australia")

ggplot(data = co2_3) +
  geom_point(mapping = aes(x = time, y = co2, color = name)) +
  xlab("Time") + ylab("CO2 Emissions")
```

- China have the greatest increase in total CO2 emission. India also have obvious increase trend. Australia have the lowest total emission in these years among these countries. USA have roughly increase trend, but recent decades were increase in decrease speed, even decrease.

#### 3(3).
```{r}
co2PC_3 <- gapminder %>%
  filter(!is.na(co2_PC)) %>% 
  group_by(time, name) %>% 
  select(name, time, co2_PC) %>%
  filter(name == "United States of America" | name == "China" | name == "India"| name == "Japan" | name == "Australia")

ggplot(data = co2PC_3) +
  geom_point(mapping = aes(x = time, y = co2_PC, color = name)) +
  xlab("Time") + ylab("CO2 Emissions Per Capita")
```

- The USA has the most significant CO2 emissions per capita. And Australia has the second largest emission, Even though Australia did not have substantial CO2 emissions in total. For China, per capita data is much less than the total emission. This figure suggests that the total CO2 emission is totally different when a change to per capita.

#### 3(4).
```{r}
avgcontinent <- gapminder %>%
  filter(time == 1960 | time == 2016) %>% 
  filter(!is.na(co2_PC)) %>% 
  filter(!is.na(region)) %>% 
  group_by(region, time) %>%
  summarize(co2mission = mean(co2_PC, na.rm = TRUE))
avgcontinent
```

- Africa, Asia and Oceania have roughly increase trend from these years. While America and Europe have decrease trend. 

#### 3(5).
```{r}
avgcontinent %>% 
  filter(time == 1960 | time == 2016) %>% 
  ggplot() +
  geom_col(aes(x = region, y = co2mission, fill = factor(time)),
           pos = "dodge") +
  xlab("Continent") + ylab("Average CO2 emissions per capita")
```

#### 3(6).
```{r}
gapminder %>% 
  filter(!is.na(name)) %>% 
  filter(time == 2016) %>%
  group_by(region) %>% 
  select(name, region, co2_PC) %>%
  arrange(co2_PC) %>%
  filter(rank(co2_PC) <= 3 |rank(desc(co2_PC)) <= 3) %>%
  arrange(region) %>% 
  print()
```

### 4.GDP per capita
#### 4(1).
```{r}
gapminder %>%
  filter(time == 1960) %>%
  filter(!is.na(GDP_PC)) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
ggplot() +
  geom_point(mapping = aes(x = GDP_PC, y = lifeExpectancy, color = region)) +
  xlab("GDP per capita") + ylab("Life expectancy")
```

- Europe have the greatest GDP per capita and life expectancy in average. Some Americas countries have both high GDP and life expectancy while most countries in Americas only have high life expectancy. Africa and Asia have relative low in life expectancy and GDP per capita. 

#### 4(2).
```{r}
gapminder %>%
  filter(!is.na(region)) %>% 
  filter(time == 2019) %>%
  filter(!is.na(GDP_PC)) %>% 
  filter(!is.na(lifeExpectancy)) %>% 
ggplot() +
  geom_point(mapping = aes(x = GDP_PC, y = lifeExpectancy, color = region)) +  xlab("GDP per capita") + ylab("Life expectancy")
```

#### 4(3).

- As the world develops, more countries are growing and progressing in life expectancy and GDP per capita. Overall, the world's life expectancy is also rising. An increasing number of Asian countries have higher GDP per capita, but most African countries are not growing in GDP per capita. Over the past 60 years, the life expectancy of all nations has increased, as has the GDP, but hopefully, Africa and some countries in the Americas will see more growth in GDP per capita in the future.

#### 4(4).
```{r}
gapminder %>%
  filter(!is.na(region)) %>% 
  group_by(region, time) %>% 
  filter(time == 1960 | time == 2019) %>% 
  summarize(avg_life = mean(lifeExpectancy, na.rm = TRUE))
```

- Yes, the results fit with what do you see on the figures. When compare average life expectancy in 1960 and 2019, every region shown greater increase. 

#### 4(5).
```{r}
gapminder %>%
  filter(time == 1960 | time == 2019) %>% 
  filter(!is.na(region)) %>% 
  group_by(region, time) %>% 
  summarize(avg_life = mean(lifeExpectancy, na.rm = TRUE)) %>% 
  mutate(growth = avg_life - lag(avg_life)) %>% 
  group_by(region) %>% 
  summarize(growth_rate = mean(growth, na.rm = TRUE)) %>% 
  arrange(desc(growth_rate))
```

- Asia and Africa have greatest growth rate. While Europe have the lowest growth rate.

#### 4(6).
```{r}
gapminder %>% 
  filter(!is.na(region)) %>% 
  filter(!is.na(GDP_PC)) %>% 
  filter(time == 1960 | time == 2019) %>% 
  ggplot(aes(x = GDP_PC, fill = factor(time))) +
  geom_histogram() +
  labs(x = "GDP per capita", y = "Countries", fill = "Time")
```

#### 4(7).
```{r}
rank_1960 <- gapminder %>% 
  filter(time == 1960) %>%
  filter(!is.na(iso2)) %>% 
  group_by(time) %>% 
  select("iso2", "time", "lifeExpectancy") %>%
  arrange(desc(lifeExpectancy)) %>% 
  mutate(rank = rank(desc(lifeExpectancy)))
rank_1960[rank_1960$iso2 == "US", ]
rank_2019 <- gapminder %>% 
  filter(time == 2019) %>%
  filter(!is.na(iso2)) %>% 
  group_by(time) %>% 
  select("iso2", "time", "lifeExpectancy") %>%
  arrange(desc(lifeExpectancy)) %>% 
  mutate(rank = rank(desc(lifeExpectancy)))
rank_2019[rank_2019$iso2 == "US", ]
```

#### 4(8).
```{r}
gapminder %>%
  filter(!is.na(region)) %>% 
  filter(time == 1960 | time == 2019) %>% 
  filter(!is.na(lifeExpectancy)) %>%
  group_by(time) %>%
  mutate(ranks = (rank(desc(lifeExpectancy)) / n_distinct(iso2))) %>% 
  filter(iso2 == "US") %>% 
  select(iso2, ranks, lifeExpectancy, time)
```

